Should geophysicists use gravity anomaly or gravity disturbance?

### 2018-05-06 manuscript/letter.tex
small change conclusion, deletion unused files
Val's general review

\title[Theoretical aspects in gravity modelling]
      {Theoretical aspects in gravity modelling}

\author[Oliveira Jr et al.]
{Vanderlei C. Oliveira Jr$^1$\thanks{Email: vanderlei@on.br}, 
Kristoffer A. T. Hallam$^{1}$, Val\'{e}ria C. F. Barbosa$^1$
and Leonardo Uieda$^2$ \\
$^1$ Department of Geophysics, Observat\'{o}rio Nacional, Rio de Janeiro, Brazil \\
$^2$ Department of Geology and Geophysics, University of Hawaii, Manoa, USA
}

@@

The first step of the procedure for isolating this 
gravitational component consists in removing the non-gravitational 
effects due to the vehicle motion and also the time variations 
such as Earth tides, instrumental drift and barometric 
pressure changes, for example.
If these effects are properly removed, the resultant 
gravity data can be considered as the sum of a 
centrifugal component due to the Earth's rotation and
a gravitational component produced by the whole Earth's
internal density distribution.
The isolation of this particular gravitational component 
and its subsequent use for estimating density 
distributions related to geological structures in subsurface 
are the main goals in applied geophysics \citep{blakely1996}.
The computation of the gravitational effect produced by
the geological structures is commonly known in geophysics as 
\textit{gravity modelling}.
This term is different from \textit{gravity field modelling}, which
is commonly used in geodesy to denote the characterization of the 
gravity field in a local, regional or global scale.

Based on well-established concepts of the literature,
we present a discussion aiming at bringing some
light to the following question: in geophysical applications,
should we use the gravity disturbance or gravity anomaly?
It seems that this theoretical issue has been 
debated within the scientific community from a 
more geodetic than geophysical point of view
\citep{lafehr1991,chapin1996,li2001,fairhead2003,
hackney-featherstone2003,hinze2005}.
Our reasoning suggests that the gravity 
disturbance is more appropriated than the gravity anomaly 
for geophysical purposes.

@@

Traditionally, the Earth's gravity field is approximated 
by the gravity field produced by a reference ellipsoid
(or level ellipsoid), which is rigid and geocentric.
This reference ellipsoid has the minor axis $b$ 
coincident with the mean rotating axis of the Earth $Z$, the 
same total mass (including the atmosphere) and also the
same angular velocity of the Earth \citep{heiskanen-moritz1967,
vanicek1987,hofmann-wellenhof-moritz2005,torge2012}.
Another characteristic of this model is that its
limiting surface coincides with a particular equipotential 
of its own gravity field.
Here, we follow \citep{torge2012} and call this model as
\textit{normal Earth}.

Similarly to the gravity vector and gravity, 
the resultant of the virtual 
gravitational and centrifugal forces exerted by the normal
Earth on a body at rest at a point $P$ is called 
\textit{normal gravity vector} and its intensity is called 
\textit{normal gravity}.
In geodesy, any model used to represent the normal gravity field
can be arbitrarily defined for the only purpose
of keeping the difference from the actual gravity field as small 
as possible \citep{vanicek1987}.

@@

In a local or regional study, geophysicists commonly use a topocentric 
Cartesian coordinate system (TCS) with origin at a point $P$ on or 
close to the Earth's surface and axes $x$, $y$ and $z$ (Fig. 
\ref{fig:TCS}). In the TCS, the axes $x$ and $y$ are parallel to 
the unit vectors $\hat{\mitbf{v}}_{P}$ $\hat{\mitbf{w}}_{P}$, respectively,
whereas the $z$-axis is opposite to the unit vector 
$\hat{\mitbf{u}}_{P}$ (eq. \ref{eq:unit-vectors}) and points downward.
Consider, for example, a TCS (Fig. \ref{fig:TCS}) with origin at a point $P$ with
coordinates $(X_{P}, Y_{P}, Z_{P})$ referred to the GCS (Fig. \ref{fig:GCS-GGS}).
In this case, the relationship between the coordinates $(X, Y, Z)$ of a 
point in the GCS and the coordinates $(x, y, z)$ of the same point in the 
TCS is given by:
\begin{equation}
\begin{bmatrix}
x \\
y \\
z 
\end{bmatrix} =
\mitbf{R}^{\top} \begin{bmatrix}
X - X_{P} \\
Y - Y_{P} \\
Z - Z_{P}
\end{bmatrix} \: ,
\label{eq:GCS2TCS}
\end{equation}
where $\mitbf{R}$ is an orthogonal matrix whose columns
are defined by the unit vectors $\hat{\mitbf{u}}$, $\hat{\mitbf{v}}$ and
$\hat{\mitbf{w}}$ (eq. \ref{eq:unit-vectors}), evaluated at the
point $P$, as follows:
\begin{equation}
\mitbf{R} = 
\begin{bmatrix}
\hat{\mitbf{v}} & \hat{\mitbf{w}} & -\hat{\mitbf{u}}
\end{bmatrix} \: .
\label{eq:R}
\end{equation}

@@

Let $\mitbf{g}_{P}$ and $\mitbf{\gamma}_{P}$ be, respectively, the 
gravity vector (corrected from non-gravitational effects due to vehicle 
motion and time variations such as Earth tides, instrumental drift and 
barometric pressure changes, for instance) and the normal gravity vector 
at a point $P$. 
In this case, the gravity vector $\mitbf{g}_{P}$ represents the 
gradient of a scalar potential called \textit{gravity potential}, 
which is the sum of a scalar gravitational potential and a 
scalar centrifugal potential.
Similarly, the normal gravity vector $\mitbf{\gamma}_{P}$ represents the 
gradient of a scalar potential called \textit{normal potential}, 
which is also the sum of a scalar gravitational potential and a 
scalar centrifugal potential.
By definition, the centrifugal part of the normal potential is equal 
to that of the gravity potential.
Then, the both gravity vector and normal gravity vectors are represented 
by the sum of two vectors: the gradient of a gravitational potential and
the gradient of a centrifugal potential. Notice that, as the centrifugal 
part of the gravity potential and normal potential are the same, the
centrifugal part of the gravity vector and normal gravity vector are
also equal to each other.

The difference between $\mitbf{g}_{P}$ and $\mitbf{\gamma}_{P}$, at the same 
point $P$, defines a quantity called \textit{gravity disturbance 
vector}, which is given by:
\begin{equation}
\delta \mitbf{g}_{P} = 
\mitbf{g}_{P} - \mitbf{\gamma}_{P} \: .
\label{eq:gravity-disturbance-vector}
\end{equation}
As the centrifugal part of the normal gravity vector 
is equal to the centrifugal component of the gravity vector, 
the gravity disturbance vector $\delta \mitbf{g}_{P}$ represents 
a purely gravitational (and consequently harmonic) quantity, which is 
caused by contrasts between the actual internal 
density distribution of the Earth and the (unknown) internal density 
distribution of the normal Earth.
In applied geophysics, these density differences are generally 
called \textit{anomalous masses} (e.g., \citeauthor{hammer1945}, 
\citeyear{hammer1945}; \citeauthor{lafehr1965}, \citeyear{lafehr1965}),
\textit{density anomalies} (e.g., \citeauthor{forsberg1984}, 
\citeyear{forsberg1984}) or \textit{gravity sources} (e.g., 
\citeauthor{blakely1996}, \citeyear{blakely1996}). Here, we opted for 
using the last term.
Fig. \ref{fig:surfaces} illustrates the gravity vector 
$\mitbf{g}_{P}$, normal gravity vector $\mitbf{\gamma}_{P}$ and 
the gravity disturbance vector $\delta \mitbf{g}_{P}$ at a point $P$ 
located on the Earth's surface.

The difference between gravity $g_{P}$ (magnitude of the gravity 
vector $\mitbf{g}_{P}$) and normal gravity $\gamma_{P}$ 
(magnitude of the normal gravity vector $\mitbf{\gamma}_{P}$), at the 
same point $P$, is called \textit{gravity disturbance} 
\citep{heiskanen-moritz1967, hofmann-wellenhof-moritz2005} 
and can be represented as follows:
\begin{equation}
\delta g_{P} = g_{P} - \gamma_{P} \: ,
\label{eq:gravity-disturbance}
\end{equation}
where $g_{P} = \| \mitbf{g}_{P} \|$ and 
$\gamma_{P} = \| \mitbf{\gamma}_{P} \|$ are,
respectively, the gravity and the normal gravity at the
point $P$.
Notice that the gravity disturbance $\delta g_{P}$ is not equivalent 
to the magnitude of the gravity disturbance vector 
$\delta \mitbf{g}_{P}$ \citep{barthelmes2013, sanso_sideris2013}.
As properly pointed out by \citet{hackney-featherstone2003},
the gravity disturbance is a very-well established quantity in geodesy,
but appears to be less well known in geophysics.

Another anomalous quantity is the \textit{gravity anomaly vector}.
It is defined, at point $P$, as the difference between the gravity
vector at a point $Q^{\prime}$ on the Geoid (a particular
equipotential surface of the gravity potential) and the normal
gravity vector at a point $Q$ on the ellipsoid surface, both at the
same geodetic latitude and longitude (Fig. \ref{fig:surfaces}) as in
\begin{equation}
\Delta \mitbf{g}_{P} = \mitbf{g}_{Q^{\prime}} - \mitbf{\gamma}_{Q} .
\label{eq:gravity-anomaly-vector}
\end{equation}
Similarly to the gravity disturbance expression (eq.
\ref{eq:gravity-disturbance}), the gravity anomaly is given by
\begin{equation}
\Delta g_{P} = g_{Q^{\prime}} - \gamma_{Q} ,
\label{eq:gravity-anomaly}
\end{equation}
where $g_{Q^{\prime} } = \| \mitbf{g}_{Q^{\prime}} \|$ is
the magnitude of the gravity vector $\mitbf{g}_{Q}$ at $Q^{\prime}$, 
on the Geoid, and $\gamma_{Q} = \| \mitbf{\gamma}_{Q} \|$ is the 
magnitude of the normal gravity vector $\mitbf{\gamma}_{Q}$ at $Q$, 
on the reference ellipsoid (Fig. \ref{fig:surfaces}).
Notice that, by definition, the gravity anomaly depends on 
longitude and latitude only and is not a function of height 
\citep{barthelmes2013}. As a consequence, it is not possible to compute,
for example, the upward continuation of a gravity anomaly.
However, many authors in the literature have computed the upward 
continuation of gravity anomalies. All these authors have
implicitly considered that the gravity anomaly is an approximation of 
the gravity disturbance, which in turn can be represented by a 
harmonic function that is allowed to be upward continued.

Different gravity anomalies can be calculated (e.g., Free-air and 
Bouguer anomalies), depending on the corrections applied to 
them. These corrections are usually called gravity reductions. 
The Free-air anomaly, for example, is defined as follows 
\citep{blakely1996, hofmann-wellenhof-moritz2005}:
\begin{equation}
\Delta g_{P}^{a} 
= g_{P} - \left( \gamma_{Q} + \frac{\partial \gamma}{\partial h} \, H \right)
\: ,
\label{eq:free-air-anomaly}
\end{equation}
where $\frac{\partial \gamma}{\partial h} \approx -0.3086$ mGal/m is the
derivative of the normal gravity with respect to the geometric height $h$
and $H$ is the orthometric height (Fig. \ref{fig:surfaces}). 
Notice that the term between parenthesis represents an 
approximation of the normal gravity at the point $Q^{\prime}$ on the
Geoid. By using a similar approach, we can approximate the normal gravity
$\gamma_{P}$, at a point $P$ on the Earth's surface, so that the 
gravity disturbance $\delta g_{P}$ (eq. \ref{eq:gravity-disturbance}) 
can be rewritten as follows:
\begin{equation}
\delta g_{P} \approx g_{P} - 
\left( \gamma_{Q} + \frac{\partial \gamma}{\partial h} h \right) \: .
\label{eq:gravity-disturbance-approx}
\end{equation}
By inspection, one easily concludes that the absolute difference 
between the approximated gravity disturbance $\delta g_{P}$ 
(eq. \ref{eq:gravity-disturbance-approx}), at a point $P$ on or close to
the Earth's surface, and the Free-air anomaly
$\Delta g_{P}^{a}$ (eq. \ref{eq:free-air-anomaly}), is given by:
\begin{equation}
\vert \delta g_{P} - \Delta g^{a}_{P} \vert \approx 
\vert \frac{\partial \gamma}{\partial h} N \vert \: ,
\label{eq:difference-disturbance-free-air}
\end{equation}
where the geoidal undulation $N \approx h - H$ (Fig. \ref{fig:surfaces}) assumes 
that the Geoid and the surface of the reference ellipsoid 
are approximately parallel at $P$ and the surrounding area.
We know empirically that the geoidal undulation $N$ in the world 
is $\approx \pm 1$ on the oceans and reaches a 
maximum absolute value of $\approx 120$ m \citep[e.g.,][]{torge2012, 
sanso_sideris2013}. In Brazil, for example, the geoidal undulation 
reaches $\approx \pm 30$ m \citep{ibge_mapgeo2015}. Consequently, the
maximum absolute differences between gravity disturbance and
Free-air anomaly, according to eq. \ref{eq:difference-disturbance-free-air},
is $\approx 9.258$ mGal.

The approximations defined by eqs \ref{eq:gravity-disturbance-approx}
and \ref{eq:difference-disturbance-free-air} are commonly used
in geodesy to define the \textit{fundamental equation of physical
geodesy} \citep{hofmann-wellenhof-moritz2005}.
A similar approach can be used to show that the Bouguer anomaly
is an approximation of the terrain corrected gravity disturbance.
The Bouguer anomaly is commonly used by geophysicists as the
gravitational effect produced by the gravity sources.
Although this approximation is valid for most practical applications,
it is important to bear in mind not only the terminology 
changes, but also the conceptual assumptions.

@@

From eq. \ref{eq:gravity-disturbance-vector}, the gravity vector $\mitbf{g}_{i}$, at a point 
$(x_{i}, y_{i}, z_{i})$ in the TCS (Fig. \ref{fig:TCS}b), can be represented by:
\begin{equation}
\mitbf{g}_{i} = \mitbf{\gamma}_{i} + \delta \mitbf{g}_{i} \: ,
\label{eq:gravity-vector-TCS}
\end{equation}
where $\mitbf{\gamma}_{i}$ and $\delta \mitbf{g}_{i}$
are, respectively, the normal gravity vector and the 
gravity disturbance vector produced by the anomalous 
masses at the point $(x_{i}, y_{i}, z_{i})$.

The gravity $g_{i}$ can be approximated by a first order Taylor's 
expansion as follows \citep{sanso_sideris2013}:
\begin{equation}
g_{i} \approx \gamma_{i} + 
\hat{\mitbf{\gamma}}_{i}^{\top} \delta \mitbf{g}_{i} \: ,
\label{eq:gobs-approx}
\end{equation}
where $^{\top}$ denotes transposition,
$\hat{\mitbf{\gamma}}_{i} = -\hat{\mitbf{u}}_{i}$ is a unit 
vector with the same direction as the normal gravity vector 
$\mitbf{\gamma}_{i}$ at the point $(x_{i}, y_{i}, z_{i})$, in the
topocentric Cartesian coordinate system (Fig. \ref{fig:TCS}b), and
$\hat{\mitbf{u}}_{i}$ is the unit vector $\hat{\mitbf{u}}$ 
(eq. \ref{eq:unit-vectors}) evaluated at the corresponding 
point $(h_{i}, \varphi_{i}, \lambda_{i})$ in the 
GGS (Fig. \ref{fig:GCS-GGS}).
This approximation can be made because, fortunately, 
the condition $\gamma_{i} \gg \| \delta \mathbf{g}_{i} \|$ 
is met at all points located above or on the Earth's surface.
This approximation, which is known in geodesy \citep[e.g.,][]{sanso_sideris2013},
is largely used in applied geophysics for representing
total-field anomalies \citep[e.g.,][]{blakely1996}.
Notice that, in local- or regional-gravity studies, the unit
vector $\hat{\mitbf{\gamma}}_{i}$ (eq. \ref{eq:gobs-approx}) 
may be considered constant throughout the study area and 
parallel to the $z$ axis of the TCS (Fig. \ref{fig:TCS}b).
By using the approximation defined in eq. 
\ref{eq:gobs-approx}, the gravity disturbance 
(eq. \ref{eq:gravity-disturbance}) can be rewritten as 
follows 
\begin{equation}
\delta g_{i} \approx \hat{\mitbf{\gamma}}_{P}^{\top} \delta \mitbf{g}_{i} \: ,
\label{eq:gravity-disturbance-approx-TCS}
\end{equation}
where $\hat{\mitbf{\gamma}}_{P}$ represents the unit vector with
the same direction as the normal gravity vector $\mitbf{\gamma}_{P}$
at the origin $P$ of the TCS (Fig. \ref{fig:TCS}b).
This equation shows that the gravity disturbance $\delta g_{i}$ (eq. \ref{eq:gravity-disturbance}) is different from the magnitude
of the gravity disturbance vector $\delta \mitbf{g}_{i}$
produced by the gravity sources. Rather, it represents the 
component of $\delta \mitbf{g}_{i}$ on the direction of the normal
gravity vector \citep{hofmann-wellenhof-moritz2005, sanso_sideris2013}.

>>>>

\title[Gravity anomaly or gravity disturbance?]
      {Should geophysicists use gravity anomaly or gravity disturbance?}

\author[Hallam et al.]
{Kristoffer A. T. Hallam$^{1}$\thanks{Emails: kristoffer.hallam@gmail.com,
vanderlei@on.br}, Vanderlei C. Oliveira Jr$^1$,
Val\'{e}ria C. F. Barbosa$^1$ \linebreak and Leonardo Uieda$^2$ \\
$^1$ Department of Geophysics, Observat\'{o}rio Nacional, Rio de Janeiro, Brazil \\
$^2$ Department of Geology and Geophysics, University of Hawaii, Manoa, USA
}

@@

For isolating this 
gravitational component consists in removing the non-gravitational 
effects due to the vehicle motion and also the time variations 
such as Earth tides, instrumental drift and barometric 
pressure changes, for example.
If these effects are properly removed, the resultant 
gravity data can be considered as the sum of a 
centrifugal component due to the Earth's rotation and
a gravitational component produced by the whole Earth's
internal density distribution.
The isolation of this particular gravitational component 
and its subsequent use for estimating density 
distributions related to geological structures in subsurface 
are the main goals in applied geophysics \citep{blakely1996}.
The computation of the gravitational effect produced by
the geological structures is commonly known in geophysics as 
\textit{gravity modelling}.
This term is different from \textit{gravity field modelling}, which
is commonly used in geodesy to denote the characterization of the 
gravity field in a local, regional or global scales.

Based on well-established concepts of the literature,
we present a discussion aiming at critically examining
the following question: in geophysical applications,
should we use the gravity disturbance or gravity anomaly?
It seems that this theoretical issue has been 
debated within the scientific community from a 
more geodetic than geophysical point of view
\citep{lafehr1991,chapin1996,li2001,fairhead2003,
hackney-featherstone2003,hinze2005}.
Our reasoning suggests that the gravity 
disturbance is more appropriated than the gravity anomaly 
for geophysical purposes.

@@

Traditionally, the Earth's gravity field is approximated 
by the gravity field produced by a reference ellipsoid
(or level ellipsoid), which is a rigid and geocentric model.
This reference ellipsoid has the minor axis $b$ 
coincident with the mean rotating axis of the Earth $Z$, the 
same total mass (including the atmosphere) and also the
same angular velocity of the Earth \citep{heiskanen-moritz1967,
vanicek1987,hofmann-wellenhof-moritz2005,torge2012}.
Another characteristic of this model is that its
limiting surface coincides with a particular equipotential 
of its own gravity field.
Here, we follow \citep{torge2012} and call this model as
\textit{normal Earth}.

The resultant of the virtual 
gravitational and centrifugal forces exerted by the normal
Earth on a body at rest at a point $P$ is called 
\textit{normal gravity vector} and its intensity is called 
\textit{normal gravity}.
In geodesy, any model used to represent the normal gravity field
can be arbitrarily defined for the only purpose
of keeping the difference from the actual gravity field as small 
as possible \citep{vanicek1987}.

@@

In a local or regional study, geophysicists commonly use a topocentric 
Cartesian coordinate system (TCS) with origin at a point $P$ on or 
close to the Earth's surface and axes $x$, $y$ and $z$ (Fig. 
\ref{fig:TCS}). In the TCS, the axes $x$ and $y$ are parallel to 
the unit vectors $\hat{\mitbf{v}}_{P}$ $\hat{\mitbf{w}}_{P}$, respectively,
whereas the $z$-axis is opposite to the unit vector 
$\hat{\mitbf{u}}_{P}$ (eq. \ref{eq:unit-vectors}) and points downward.
Consider, for example, a TCS (Fig. \ref{fig:TCS}) with origin at a point $P$ with
coordinates $(X_{P}, Y_{P}, Z_{P})$ referred to the GCS (Fig. \ref{fig:GCS-GGS}).
In this case, the relationship between the coordinates $(X, Y, Z)$ of a 
point in the GCS and the coordinates $(x, y, z)$ of the same point in the 
TCS is given by:
\begin{equation}
\begin{bmatrix}
x \\
y \\
z 
\end{bmatrix} =
\mitbf{R}^{\top} \begin{bmatrix}
X - X_{P} \\
Y - Y_{P} \\
Z - Z_{P}
\end{bmatrix} \: ,
\label{eq:GCS2TCS}
\end{equation}
where $\mitbf{R}$ is a $3 \times 3$ orthogonal matrix whose 
first, second and third columns are defined, respectively, 
by the unit vectors $\hat{\mitbf{v}}$, $\hat{\mitbf{w}}$ 
and $-\hat{\mitbf{u}}$ (eq. \ref{eq:unit-vectors}), 
evaluated at the point $P$.

@@

Let $\mitbf{g}_{P}$ and $\mitbf{\gamma}_{P}$ be, respectively, the 
gravity vector (corrected from non-gravitational effects due to vehicle 
motion and time variations such as Earth tides, instrumental drift and 
barometric pressure changes, for instance) and the normal gravity vector 
at a point $P$. 
In this case, the gravity vector $\mitbf{g}_{P}$ represents the 
gradient of a scalar potential called \textit{gravity potential}, 
which is the sum of a scalar gravitational potential and a 
scalar centrifugal potential.
Similarly, the normal gravity vector $\mitbf{\gamma}_{P}$ represents the 
gradient of a scalar potential called \textit{normal potential}, 
which is also the sum of a scalar gravitational potential and a 
scalar centrifugal potential.
By definition, the centrifugal part of the normal potential is equal 
to that of the gravity potential.

The difference between $\mitbf{g}_{P}$ and $\mitbf{\gamma}_{P}$, at the same 
point $P$, defines a quantity called \textit{gravity disturbance 
vector}, which is given by:
\begin{equation}
\delta \mitbf{g}_{P} = 
\mitbf{g}_{P} - \mitbf{\gamma}_{P} \: .
\label{eq:gravity-disturbance-vector}
\end{equation}
Because the centrifugal part of the normal gravity vector 
is equal to the centrifugal component of the gravity vector, 
the gravity disturbance vector $\delta \mitbf{g}_{P}$ represents 
a purely gravitational (and consequently harmonic) quantity, which is 
caused by contrasts between the actual internal 
density distribution of the Earth and the internal density 
distribution of the normal Earth.
It is important to stress that both density distributions are unknown.
In applied geophysics, these density differences are generally 
called \textit{anomalous masses} (e.g., \citeauthor{hammer1945}, 
\citeyear{hammer1945}; \citeauthor{lafehr1965}, \citeyear{lafehr1965}),
\textit{density anomalies} (e.g., \citeauthor{forsberg1984}, 
\citeyear{forsberg1984}) or \textit{gravity sources} (e.g., 
\citeauthor{blakely1996}, \citeyear{blakely1996}). Here, we opted for 
using the last term.
Fig. \ref{fig:surfaces} illustrates the gravity vector 
$\mitbf{g}_{P}$, normal gravity vector $\mitbf{\gamma}_{P}$ and 
the gravity disturbance vector $\delta \mitbf{g}_{P}$ at a point $P$ 
located on the Earth's surface.

The difference between the magnitudes of gravity vector
$g_{P} = \| \mitbf{g}_{P} \|$ and the normal gravity vector
$\gamma_{P} = \| \mitbf{\gamma}_{P} \|$, at the 
same point $P$, is called \textit{gravity disturbance} 
\citep{heiskanen-moritz1967, hofmann-wellenhof-moritz2005} 
and can be represented as follows:
\begin{equation}
\delta g_{P} = g_{P} - \gamma_{P} \: .
\label{eq:gravity-disturbance}
\end{equation}
Notice that the gravity disturbance $\delta g_{P}$ is not equivalent 
to the magnitude of the gravity disturbance vector 
$\delta \mitbf{g}_{P}$ \citep{barthelmes2013, sanso_sideris2013}.
Another anomalous quantity is the \textit{gravity anomaly vector}.
It is defined, at point $P$, as the difference between the gravity
vector at a point $Q^{\prime}$ on the Geoid (a particular
equipotential surface of the gravity potential) and the normal
gravity vector at a point $Q$ on the ellipsoid surface, both at the
same geodetic latitude and longitude (Fig. \ref{fig:surfaces}) as in
\begin{equation}
\Delta \mitbf{g}_{P} = \mitbf{g}_{Q^{\prime}} - \mitbf{\gamma}_{Q} .
\label{eq:gravity-anomaly-vector}
\end{equation}
Similarly to the gravity disturbance expression (eq.
\ref{eq:gravity-disturbance}), the gravity anomaly is given by
\begin{equation}
\Delta g_{P} = g_{Q^{\prime}} - \gamma_{Q} ,
\label{eq:gravity-anomaly}
\end{equation}
where $g_{Q^{\prime} } = \| \mitbf{g}_{Q^{\prime}} \|$ is
the magnitude of the gravity vector $\mitbf{g}_{Q}$ on the Geoid,
at $Q^{\prime}$, and $\gamma_{Q} = \| \mitbf{\gamma}_{Q} \|$ is the 
magnitude of the normal gravity vector $\mitbf{\gamma}_{Q}$ 
on the reference ellipsoid, at the point $Q$ (Fig. \ref{fig:surfaces}).
Notice that, by definition, the gravity anomaly depends on 
longitude and latitude only and is not a function of height 
\citep{barthelmes2013}. As a consequence, it is not possible to compute,
for example, the upward continuation of a gravity anomaly.
However, many authors in the literature have computed the upward 
continuation of gravity anomalies. All these authors have
implicitly considered that the gravity anomaly is an approximation of 
the gravity disturbance, which in turn can be represented by a 
harmonic function that is allowed to be upward continued.

Different gravity anomalies can be calculated, 
depending on the corrections applied to them. 
These corrections are usually called \textit{gravity reductions}. 
The Free-air anomaly, for example, may be defined as 
follows \citep{blakely1996, hofmann-wellenhof-moritz2005}:
\begin{equation}
\Delta g_{P}^{F} 
= g_{P} - \left( \gamma_{Q} + \frac{\partial \gamma}{\partial h} \, H_{P} \right) \: ,
\label{eq:free-air-anomaly}
\end{equation}
where $\frac{\partial \gamma}{\partial h} \approx -0.3086$ mGal/m is the
derivative of the normal gravity with respect to the geometric height $h$
and $H_{P}$ is the orthometric height $H$ (Fig. \ref{fig:surfaces}) at the
point $P$.
Notice that the term between parenthesis represents an 
approximation of the normal gravity at the point $Q^{\prime}$ on the
Geoid. By using a similar approach, we can approximate the normal gravity
$\gamma_{P}$, at a point $P$ on the Earth's surface, so that the 
gravity disturbance $\delta g_{P}$ (eq. \ref{eq:gravity-disturbance}) 
can be rewritten as follows:
\begin{equation}
\delta g_{P} \approx g_{P} - 
\left( \gamma_{Q} + \frac{\partial \gamma}{\partial h} h \right) \: .
\label{eq:gravity-disturbance-approx}
\end{equation}
By inspection, one easily concludes that the absolute difference 
between the approximated gravity disturbance $\delta g_{P}$ 
(eq. \ref{eq:gravity-disturbance-approx}), at a point $P$ on or close to
the Earth's surface, and the Free-air anomaly
$\Delta g_{P}^{F}$ (eq. \ref{eq:free-air-anomaly}), is given by:
\begin{equation}
\vert \delta g_{P} - \Delta g^{F}_{P} \vert \approx 
\vert \frac{\partial \gamma}{\partial h} N \vert \: ,
\label{eq:difference-disturbance-free-air}
\end{equation}
where $N \approx h - H$ is the geoidal undulation (Fig. \ref{fig:surfaces}).
This approximation assumes that the Geoid and the surface of 
the reference ellipsoid are approximately parallel at $P$ 
and the surrounding area.
We know empirically that the geoidal undulation $N$ in the world 
is $\approx \pm 1$ m on the oceans and reaches a 
maximum absolute value of $\approx 120$ m \citep[e.g.,][]{torge2012, 
sanso_sideris2013}. In Brazil, for example, the geoidal undulation 
reaches $\approx \pm 30$ m \citep{ibge_mapgeo2015}. Consequently, the
maximum absolute differences between gravity disturbance and
Free-air anomaly, according to eq. \ref{eq:difference-disturbance-free-air},
is $\approx 9.258$ mGal.
The approximations defined by eqs \ref{eq:gravity-disturbance-approx}
and \ref{eq:difference-disturbance-free-air} are commonly used
in geodesy to define the \textit{fundamental equation of physical
geodesy} \citep{hofmann-wellenhof-moritz2005}.

The simple Bouguer anomaly is commonly used by geophysicists as the
gravitational effect produced by the gravity sources.
It is defined, over the continents, as follows 
\citep{blakely1996, hofmann-wellenhof-moritz2005}:
\begin{equation}
\Delta g_{P}^{B} 
= \Delta g_{P}^{F} - 2 \pi G \rho_{t} \, H_{P} \: ,
\label{eq:simple-bouguer-anomaly}
\end{equation}
where $\Delta g_{P}^{F}$ is the Free-air anomaly (eq. \ref{eq:free-air-anomaly})
and the last term on the right side is the simple Bouguer correction.
This term approximates the gravitational attraction that all topographic 
masses above the Geoid exerts at the point $P$ by the gravitational 
attraction of a homogeneous, infinitely extended slab of 
constant density $\rho_{t}$ and thickness equal to $H_{P}$, which is 
the orthometric height $H$ (Fig. \ref{fig:surfaces}) at the point $P$.
By computing the simple Bouguer correction at a set of points
of a survey around $P$, we approximately remove, 
from the gravity measured at the points of the survey,
the gravitational effect of a homogeneous model representing 
the topographic masses above the Geoid.
It is evident from eqs \ref{eq:free-air-anomaly}, \ref{eq:gravity-disturbance-approx}
and \ref{eq:difference-disturbance-free-air} that the simple Bouguer
anomaly $\Delta g_{P}^{B}$ (eq. \ref{eq:simple-bouguer-anomaly}) represents
the gravity disturbance $\delta g_{P}$ (eq. \ref{eq:gravity-disturbance})
minus the gravitational effect produced by this homogeneous topographic model.
Although this approximation is valid for most practical applications,
it is important to bear in mind not only the terminology 
changes, but also the conceptual assumptions.

@@

From eq. \ref{eq:gravity-disturbance-vector}, the gravity vector $\mitbf{g}_{i}$, at a point 
$(x_{i}, y_{i}, z_{i})$ in the TCS (Fig. \ref{fig:TCS}a), can be represented by:
\begin{equation}
\mitbf{g}_{i} = \mitbf{\gamma}_{i} + \delta \mitbf{g}_{i} \: ,
\label{eq:gravity-vector-TCS}
\end{equation}
where $\mitbf{\gamma}_{i}$ and $\delta \mitbf{g}_{i}$
are, respectively, the normal gravity vector and the 
gravity disturbance vector produced by the anomalous 
masses at the point $(x_{i}, y_{i}, z_{i})$.

The gravity $g_{i}$ can be approximated by a first order Taylor's 
expansion as follows \citep{sanso_sideris2013}:
\begin{equation}
g_{i} \approx \gamma_{i} + 
\hat{\mitbf{\gamma}}_{i}^{\top} \delta \mitbf{g}_{i} \: ,
\label{eq:gobs-approx}
\end{equation}
where $^{\top}$ denotes transposition,
$\hat{\mitbf{\gamma}}_{i} = -\hat{\mitbf{u}}_{i}$ is a unit 
vector with the same direction as the normal gravity vector 
$\mitbf{\gamma}_{i}$ at the point $(x_{i}, y_{i}, z_{i})$, in the
topocentric Cartesian coordinate system (Fig. \ref{fig:TCS}a), and
$\hat{\mitbf{u}}_{i}$ is the unit vector $\hat{\mitbf{u}}$ 
(eq. \ref{eq:unit-vectors}) evaluated at the corresponding 
point $(h_{i}, \varphi_{i}, \lambda_{i})$ in the 
GGS (Fig. \ref{fig:GCS-GGS}).
This approximation can be made because, fortunately, 
the condition $\gamma_{i} \gg \| \delta \mathbf{g}_{i} \|$ 
is met at all points located above or on the Earth's surface.
This approximation is known in geodesy \citep[e.g.,][]{sanso_sideris2013}
and a similar approximation is largely used in applied geophysics 
for representing total-field anomalies \citep[e.g.,][]{blakely1996}.
Notice that, in local- or regional-gravity studies, the unit
vector $\hat{\mitbf{\gamma}}_{i}$ (eq. \ref{eq:gobs-approx}) 
may be considered constant throughout the study area and 
parallel to the $z$ axis of the TCS (Fig. \ref{fig:TCS}a).
By using the approximation defined in eq. 
\ref{eq:gobs-approx}, the gravity disturbance 
(eq. \ref{eq:gravity-disturbance}) can be rewritten as 
follows 
\begin{equation}
\delta g_{i} \approx \hat{\mitbf{\gamma}}_{P}^{\top} \delta \mitbf{g}_{i} \: ,
\label{eq:gravity-disturbance-approx-TCS}
\end{equation}
where $\hat{\mitbf{\gamma}}_{P}$ represents the unit vector with
the same direction as the normal gravity vector $\mitbf{\gamma}_{P}$
at the origin $P$ of the TCS (Fig. \ref{fig:TCS}a).
This equation shows that the gravity disturbance $\delta g_{i}$ (eq. \ref{eq:gravity-disturbance}) is different from the magnitude
of the gravity disturbance vector $\delta \mitbf{g}_{i}$
produced by the gravity sources. Rather, it represents the 
component of $\delta \mitbf{g}_{i}$ on the direction of the normal
gravity vector \citep{hofmann-wellenhof-moritz2005, sanso_sideris2013}.
